Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-16 manuscript.qmd

![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.svg){#fig-data-strat}

@@

![Monitoring quantities during the burn-in phase of the parameter sampling. Left: The autocorrelation time of the sampling chain estimated at different sampling steps. We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps. Right: Fraction of accepted MCMC proposals averaged over all parallel walkers. Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}

@@

![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}

@@

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.svg){#fig-model-prevalences-midext width="50%"}

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig-model-risks.svg){#fig-model-risks}

>>>>

![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig_data_strat.svg){#fig-data-strat}

@@

![Monitoring quantities during the burn-in phase of the parameter sampling. Left: The autocorrelation time of the sampling chain estimated at different sampling steps. We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps. Right: Fraction of accepted MCMC proposals averaged over all parallel walkers. Values around 30% indicate good mixing of the walkers.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}

@@

![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}

@@

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
